Tempo and mode of promoter evolution as observed in the \textit{Paramecium aurelia} species complex

### 2020-08-03 manuscript.tex
Added several missing references

Main et al. \citep{Main:2013dg}, used a separate TSS profiling method in \textit{D. melanogaster} to identify putative orthologous TSSs in three other \textit{Drosophila} taxa: \textit{D. simulans}, \textit{D. pseudoobscura} and \textit{D. sechellia}. The authors found that the location and estimated activities of orthologous TSSs to be generally conserved, but did observe instances of TSS variation indicative of new promoters. Notably, Main \textit{et. al.} found differences in promoter selection in the Alcohol dehydrogenase (ADH) gene across the four \textit{Drosophila} species \citep{Main:2013dg}. Working with high-resolution promoter data in human and mouse, Young and colleagues \citep{Young:2015xx} found that more than half (56\$) of tissue-matched promoters were lost, or turned over, between the two species. Noting that most protein-coding genes in mammals \citep{FANTOMConsortiumandtheRIKENPMIandCLSTDGT:2014hz} have multiple promoters, the authors propose that promoter turnover is an important but overlooked means by which transcriptional modules diverge during evolution. Li and colleagues (2018) investigated the fates of TSSs across  the primate lineage. They found that new TSSs frequently emerge in transposable elements and repetitive regions of DNA, and are more likely to emerge in areas near existing regulatory regions. While this work was valuable, demonstrating potential routes of regulatory innovation, the study was limited by the fact it only used empirically-defined TSSs in human (as those from other primates were not available), a fact that prevented the study from identifying novel promoters in the other primate taxa in the study. \\

Duplicate genes are a major source of evolutionary change (cite \citep{Ohno:1970wm} Moore and Puruggnan, 2006), and their evolutionary fates have long been a subject of intense study (reviewed in Taylor and Raes). The availability of high-quality genome assemblies and genome annotations has allowed us to investigate duplicate gene evolution in a variety of phylogenetic contexts (cite Gu et al., 2003 and others). However, while the fates of ortho-paralog pairs has been heavily studied (enter citations for \citep{Ohno:1970wm}; Wolfe and Shields, 1997, Force et al., 1999, others), the fates of \textit{cis}-regulatory regions of duplicate gene pairs is much less well-understood. Haberer \textit{et al}. (2004) analyzed the \textit{cis}-regulatory regions of both tandem- and and segmantally duplicated genes in \textit{Arabidopsis thaliana}. Using sequence alignments, Haberer and colleages found evidence for conservation in the promoters of both tandem and segmental duplicates, with the more recently diverged tandem duplicate promoters having higher sequence similarity and gene expression correlation.\\

@@

The relative behavior of duplicate gene promoters that arose through different mechanisms are of interest. Arsvoski \textit{et al}., also working in \textit{A. thaliana}, incorporated DNAse I footprinting data into their analysis, asking whether duplicates from whole-genome duplicates (WGD) exhibited differences in their network connectivity compared with those that arose through tandem duplication. (Continue here)

>>>>

Main et al. \citep{Main:2013dg}, used a separate TSS profiling method in \textit{D. melanogaster} to identify putative orthologous TSSs in three other \textit{Drosophila} taxa: \textit{D. simulans}, \textit{D. pseudoobscura} and \textit{D. sechellia}. The authors found that the location and estimated activities of orthologous TSSs to be generally conserved, but did observe instances of TSS variation indicative of new promoters. Notably, Main \textit{et. al.} found differences in promoter selection in the Alcohol dehydrogenase (ADH) gene across the four \textit{Drosophila} species \citep{Main:2013dg}. Working with high-resolution promoter data in human and mouse, Young and colleagues \citep{Young:2015xx} found that more than half (56\$) of tissue-matched promoters were lost, or turned over, between the two species. Noting that most protein-coding genes in mammals \citep{FANTOMConsortiumandtheRIKENPMIandCLSTDGT:2014hz} have multiple promoters, the authors propose that promoter turnover is an important but overlooked means by which transcriptional modules diverge during evolution. Li and colleagues (2018) \citep{Li.2018} investigated the fates of TSSs across the primate lineage. They found that new TSSs frequently emerge in transposable elements and repetitive regions of DNA, and are more likely to emerge in areas near existing regulatory regions. While this work was valuable, demonstrating potential routes of regulatory innovation, the study was limited by the fact it only used empirically-defined TSSs in human (as those from other primates were not available), a fact that prevented the study from identifying novel promoters in the other primate taxa in the study. \\

Duplicate genes are a major source of evolutionary change (cite \citep{Ohno:1970wm} Moore and Puruggnan, 2006), and their evolutionary fates have long been a subject of intense study (reviewed in \citep{Taylor.2004}). The availability of high-quality genome assemblies and genome annotations has allowed us to investigate duplicate gene evolution in a variety of phylogenetic contexts  \citep{Gu.2003} (cite others). However, while the fates of ortho-paralog pairs has been heavily studied (enter citations for \citep{Ohno:1970wm}; \citep{Wolfe.1997, Force.1999}, the fates of \textit{cis}-regulatory regions of duplicate gene pairs is much less well-understood. Haberer \textit{et al}. (2004) analyzed the \textit{cis}-regulatory regions of both tandem- and and segmantally duplicated genes in \textit{Arabidopsis thaliana}. Using sequence alignments, Haberer and colleagues found evidence for conservation in the promoters of both tandem and segmental duplicates, with the more recently diverged tandem duplicate promoters having higher sequence similarity and gene expression correlation.\\

@@

The relative behavior of duplicate gene promoters that arose through different mechanisms are of interest. Arsvoski \textit{et al}. \citep{Arsovski.2015}, also working in \textit{A. thaliana}, incorporated DNAse I footprinting data into their analysis, asking whether duplicates from whole-genome duplicates (WGD) exhibited differences in their network connectivity compared with those that arose through tandem duplication. (Continue here)
